A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
use a ~ (non-breaking space) before citations

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

Jython\cite{webjython} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

\paragraph{HTM} provides us with transactions like any TM system does.
It can be used as a direct replacement for the GIL\cite{nicholas06,odaira14,fuad10}. However, as is
common with hardware-only solutions, there are quite a few limitations
that can not be lifted easily. For this comparison, we look at the
implementation of Intel in recent Haswell generation CPUs.

HTM in these CPUs works on the level of caches. This has a few
consequences like false-sharing on the cache-line level, and most
importantly it limits the amount of memory that can be accessed within
a transaction. This transaction-length limitation makes it necessary
to have a fallback in place in case this limit is reached. In recent
attempts, the usual fallback is the GIL\cite{odaira14,fuad10}. In our
experiments, the current generation of HTM proved to be very fragile
and thus needing the fallback very often. Consequently, scalability
suffered a lot from this.

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks because that would require
much longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its
performance.  It is not unusual for the overhead introduced by STM to
be between 100\% to even 1000\% \cite{cascaval08,drago11}. While STM
systems often scale very well to a big number of threads and
eventually overtake the single-threaded execution, they often provide
no benefits at all for low numbers of threads (1-8). There are some
attempts \cite{warmhoff13,spear09} that can reduce the overhead a lot,
but scale badly or only for certain workloads. Often the benefits
on more than one thread are too little in real world applications.

@@

While one can argue that STM requires the insertion of read and write
barriers in the whole program, this can be done automatically and
locally by a program transformation\cite{felber07}. There are attempts
to do the same for fine-grained locking\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable.
The effectiveness of these approaches is doubtful in our use case,
since we execute bytecode instructions in any order defined by a
script only known at runtime. This makes it close to impossible to
order locks consistently or to know in advance which locks a
transaction will need.

>>>>

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

\paragraph{HTM} provides us with transactions like any TM system does.
It can be used as a direct replacement for the GIL~\cite{nicholas06,odaira14,fuad10}. However, as is
common with hardware-only solutions, there are quite a few limitations
that can not be lifted easily. For this comparison, we look at the
implementation of Intel in recent Haswell generation CPUs.

HTM in these CPUs works on the level of caches. This has a few
consequences like false-sharing on the cache-line level, and most
importantly it limits the amount of memory that can be accessed within
a transaction. This transaction-length limitation makes it necessary
to have a fallback in place in case this limit is reached. In recent
attempts, the usual fallback is the GIL~\cite{odaira14,fuad10}. In our
experiments, the current generation of HTM proved to be very fragile
and thus needing the fallback very often. Consequently, scalability
suffered a lot from this.

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks because that would require
much longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its
performance.  It is not unusual for the overhead introduced by STM to
be between 100\% to even 1000\% ~\cite{cascaval08,drago11}. While STM
systems often scale very well to a big number of threads and
eventually overtake the single-threaded execution, they often provide
no benefits at all for low numbers of threads (1-8). There are some
attempts ~\cite{warmhoff13,spear09} that can reduce the overhead a lot,
but scale badly or only for certain workloads. Often the benefits
on more than one thread are too little in real world applications.

@@

While one can argue that STM requires the insertion of read and write
barriers in the whole program, this can be done automatically and
locally by a program transformation~\cite{felber07}. There are attempts
to do the same for fine-grained locking~\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable.
The effectiveness of these approaches is doubtful in our use case,
since we execute bytecode instructions in any order defined by a
script only known at runtime. This makes it close to impossible to
order locks consistently or to know in advance which locks a
transaction will need.
